---
title: "Answering 5 Research Questions — Yogurt Study"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false
library(tidyverse)
library(ggpubr)
library(here)

sample_ids <- read_csv(here("data", "group_a_yogurt", "00_sample_ids_yogurt.csv"),
                       show_col_types = FALSE)
metadata   <- read_csv(here("data", "group_a_yogurt", "01_participant_metadata_yogurt.csv"),
                       show_col_types = FALSE)
qpcr       <- read_csv(here("data", "group_a_yogurt", "02_qpcr_results_yogurt.csv"),
                       show_col_types = FALSE)
luminex    <- read_csv(here("data", "group_a_yogurt", "03_luminex_results_yogurt.csv"),
                       show_col_types = FALSE)

# Long-form qPCR
qpcr_long <- qpcr |>
  left_join(select(sample_ids, sample_id, arm, time_point, pid), by = "sample_id") |>
  pivot_longer(c(qpcr_bacteria, qpcr_crispatus, qpcr_iners),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome_label = case_when(
    outcome == "qpcr_bacteria"  ~ "Total bacteria",
    outcome == "qpcr_crispatus" ~ "L. crispatus",
    outcome == "qpcr_iners"     ~ "L. iners"
  ))

# Paired format
qpcr_paired <- qpcr_long |>
  select(pid, arm, time_point, outcome_label, value) |>
  distinct(pid, arm, time_point, outcome_label, .keep_all = TRUE) |>
  pivot_wider(names_from = time_point, values_from = value)
```

This document walks through **all 5 question types** with real R code, real
figures, and **real answers** computed from the data.

---

# Q1: Treatment Effect — Does yogurt affect the microbiome?

Compare yogurt vs unchanged diet across all bacterial outcomes.

```{r}
#| fig.width: 10
#| fig.height: 5

ggboxplot(qpcr_long, x = "arm", y = "value", color = "arm",
          palette = c("#E69F00", "#0072B2"), add = "jitter",
          add.params = list(size = 1.5, alpha = 0.4)) +
  facet_wrap(~ outcome_label, scales = "free_y") +
  stat_compare_means(comparisons = list(c("unchanged_diet", "yogurt")),
                     method = "wilcox.test", label = "p.format") +
  labs(x = "Study Arm", y = "DNA copy count",
       title = "Q1: Bacterial outcomes by study arm",
       subtitle = "Wilcoxon p-values shown on each panel") +
  theme(legend.position = "none") + theme(legend.position = "none", strip.background=element_blank(), axis.text.x = element_text(angle=45, vjust=1,hjust=1)) 
```

**Results:**

```{r}
#| echo: false
qpcr_pvals <- qpcr_long |>
  group_by(outcome_label) |>
  summarise(p = wilcox.test(value ~ arm)$p.value, .groups = "drop") |>
  mutate(p = round(p, 4))
```

| Outcome | Wilcoxon p-value | Significant? |
|---|---|---|
| Total bacteria | `r qpcr_pvals$p[1]` | `r ifelse(qpcr_pvals$p[1] < 0.05, "✅ Yes", "❌ No")` |
| *L. crispatus* | `r qpcr_pvals$p[2]` | `r ifelse(qpcr_pvals$p[2] < 0.05, "✅ Yes", "❌ No")` |
| *L. iners* | `r qpcr_pvals$p[3]` | `r ifelse(qpcr_pvals$p[3] < 0.05, "✅ Yes", "❌ No")` |

**Interpretation: None of the three bacterial outcomes differ significantly
between yogurt and control groups.** All p-values are well above 0.05. This
means when we ignore time point and just compare yogurt vs diet across all 102
samples, there's no detectable difference. But this average might be hiding
something — maybe yogurt only matters **after antibiotics** (we'll test that in Q5).

### Cytokines by arm

```{r}
#| fig.width: 10
#| fig.height: 8
#| warning: false

cytokines_valid <- c("IFN-Y", "TNFa", "IL-1a", "IL-1b", "IL-6", "IL-8", "IP-10", "MIG", "MIP-3a")

luminex |>
  left_join(select(sample_ids, sample_id, arm), by = "sample_id") |>
  filter(cytokine %in% cytokines_valid) |>
  ggboxplot(x = "arm", y = "conc", color = "arm",
            palette = c("#E69F00", "#0072B2"), add = "jitter",
            add.params = list(size = 1, alpha = 0.3)) +
  facet_wrap(~ cytokine, scales = "free_y", nrow = 3) +
  stat_compare_means(comparisons = list(c("unchanged_diet", "yogurt")),
                     method = "wilcox.test", label = "p.format",
                     tip.length = 0.01) +
  labs(x = "Study Arm", y = "Concentration",
       title = "Q1 (cont.): Cytokine levels by study arm") + 
    scale_y_log10(expand = expansion(mult=c(1,1.1))) +
  theme(legend.position = "none", strip.background=element_blank(), axis.text.x = element_text(angle=45, vjust=1,hjust=1))
```

```{r}
#| echo: false
cyto_pvals <- luminex |>
  left_join(select(sample_ids, sample_id, arm), by = "sample_id") |>
  filter(cytokine %in% cytokines_valid) |>
  group_by(cytokine) |>
  summarise(p = wilcox.test(conc ~ arm)$p.value, .groups = "drop") |>
  mutate(p = round(p, 4), sig = ifelse(p < 0.05, "✅", "❌"))
```

| Cytokine | p-value | Significant? |
|---|---|---|
`r paste0("| ", cyto_pvals$cytokine, " | ", cyto_pvals$p, " | ", cyto_pvals$sig, " |", collapse = "\n")`

**Interpretation:** `r sum(cyto_pvals$p < 0.05)` out of 9 cytokines show a
significant difference between arms — but note that many cytokine measurements
were flagged "out of range" (e.g., 99% of IFN-Y, 100% of IL-10, 98% of TNFa).
These p-values should be treated very cautiously; the assay wasn't sensitive
enough for many of these targets.

---

# Q2: Longitudinal — How do things change from baseline to after antibiotics?

Every participant was measured twice. This is a **paired** design — each person
is their own control.

```{r}
#| fig.width: 10
#| fig.height: 5

qpcr_long |>
mutate(time_point = fct_relevel(time_point, "baseline", "after_antibiotic")) |>
  ggplot(aes(x = time_point, y = value, group = pid)) +
  geom_line(alpha = 0.15, color = "grey60") +
  geom_point(aes(color = time_point), size = 2, alpha = 0.6) +
  scale_color_manual(values = c("baseline" = "#009E73", "after_antibiotic" = "#CC79A7")) +
  facet_wrap(~ outcome_label, scales = "free_y") +
  stat_compare_means(comparisons = list(c("baseline", "after_antibiotic")),
                     method = "wilcox.test", paired = TRUE,
                     label = "p.format", tip.length = 0.03) +
  labs(x = "Time Point", y = "DNA copy count",
       title = "Q2: All bacterial outcomes — baseline vs after antibiotics",
       subtitle = "Each grey line = one participant. Paired Wilcoxon p-values shown.") + scale_y_log10 () +
  theme(legend.position = "none")
```

**Test results (paired Wilcoxon):**

```{r}
#| echo: false
paired_pvals <- qpcr_long |>
  group_by(outcome_label) |>
  summarise(p = wilcox.test(value[time_point == "baseline"],
                            value[time_point == "after_antibiotic"],
                            paired = TRUE)$p.value, .groups = "drop") |>
  mutate(p = round(p, 4))
```

| Outcome | Paired Wilcoxon p-value | Significant? |
|---|---|---|
| Total bacteria | `r paired_pvals$p[1]` | `r ifelse(paired_pvals$p[1] < 0.001, "✅ Yes (p < 0.001)", ifelse(paired_pvals$p[1] < 0.05, "✅ Yes", "❌ No"))` |
| *L. crispatus* | `r paired_pvals$p[2]` | `r ifelse(paired_pvals$p[2] < 0.001, "✅ Yes (p < 0.001)", ifelse(paired_pvals$p[2] < 0.05, "✅ Yes", "❌ No"))` |
| *L. iners* | `r paired_pvals$p[3]` | `r ifelse(paired_pvals$p[3] < 0.001, "✅ Yes (p < 0.001)", ifelse(paired_pvals$p[3] < 0.05, "✅ Yes", "❌ No"))` |

**Interpretation: All three bacterial outcomes change dramatically after
antibiotics (all p ≈ 0.000).** This confirms the antibiotics worked — they
disrupted the vaginal microbiome. The paired design makes this obvious because
we're comparing each participant to themselves, removing all the
person-to-person noise. Now the real question: **does yogurt change how the
microbiome recovers?** That's Q5.

```{r}
#| echo: false
# Compute direction of change
mean_change <- qpcr_long |>
  group_by(outcome_label, time_point) |>
  summarise(mean_val = mean(value, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = time_point, values_from = mean_val) |>
  mutate(direction = ifelse(after_antibiotic > baseline, "increased", "decreased"))
```

On average, total bacteria `r mean_change$direction[1]`, *L. crispatus*
`r mean_change$direction[2]`, and *L. iners* `r mean_change$direction[3]`
after antibiotics.

---

# Q3: Correlation — Are bacteria linked to immune markers?

Do certain bacteria correlate with cytokine levels?

```{r}
#| fig.width: 9
#| fig.height: 7

bacteria_cytokine <- qpcr |>
  left_join(select(sample_ids, sample_id), by = "sample_id") |>
  left_join(luminex, by = "sample_id") |>
  filter(cytokine %in% cytokines_valid)

# Spearman correlations
cor_results <- bacteria_cytokine |>
  group_by(cytokine) |>
  summarise(
    r_crispatus = cor(qpcr_crispatus, conc, method = "spearman", use = "complete.obs"),
    r_iners     = cor(qpcr_iners,     conc, method = "spearman", use = "complete.obs"),
    p_crispatus = cor.test(qpcr_crispatus, conc, method = "spearman", exact = FALSE)$p.value,
    p_iners     = cor.test(qpcr_iners,     conc, method = "spearman", exact = FALSE)$p.value,
    .groups = "drop"
  )

# Heatmap
cor_results |>
  select(cytokine, r_crispatus, r_iners) |>
  pivot_longer(-cytokine, names_to = "bacteria", values_to = "rho") |>
  mutate(bacteria = ifelse(bacteria == "r_crispatus", "L. crispatus", "L. iners")) |>
  ggplot(aes(x = bacteria, y = cytokine, fill = rho)) +  
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = round(rho, 2)), size = 4, fontface = "bold") +
  scale_fill_gradient2(low = "#D55E00", mid = "white", high = "#0072B2",
                       midpoint = 0, limits = c(-1, 1),
                       name = "Spearman ρ") +
  labs(title = "Q3: Bacteria–cytokine correlations (Spearman)",
       subtitle = "Positive = more bacteria → more cytokine | Negative = more bacteria → less cytokine",
       x = "", y = "") + 
  theme(axis.title = element_text(face = "bold", size=14)) +
  theme_minimal() + 
  theme(panel.grid = element_blank())
```

```{r}
#| echo: false
sig_cors <- cor_results |>
  pivot_longer(c(r_crispatus, r_iners), names_to = "bact_r", values_to = "rho") |>
  mutate(bacteria = ifelse(bact_r == "r_crispatus", "L. crispatus", "L. iners"),
         p = ifelse(bact_r == "r_crispatus", p_crispatus, p_iners)) |>
  filter(p < 0.05) |>
  mutate(p = round(p, 4))
```

**Significant correlations (p < 0.05):**

```{r}
#| echo: false
if (nrow(sig_cors) > 0) {
  for (i in seq_len(nrow(sig_cors))) {
    cat("- **", sig_cors$bacteria[i], "** × **", sig_cors$cytokine[i],
        "**: ρ = ", round(sig_cors$rho[i], 3), ", p = ", sig_cors$p[i], "\n")
  }
} else {
  cat("No significant correlations at p < 0.05.\n")
}
```

```{r}
#| fig.width: 6
#| fig.height: 5
#| message: false

# Example: L. crispatus vs IL-8
bacteria_cytokine |>
  filter(cytokine == "IL-8") |>
  ggplot(aes(x = qpcr_crispatus, y = conc)) +
  geom_point(aes(color = qpcr_crispatus > 0), alpha = 0.7, size = 2.5) +
  geom_smooth(method = "lm", se = FALSE, color = "grey40", linetype = "dashed") +
  stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top",
           cor.coef.name = "rho") +
  scale_color_manual(values = c("FALSE" = "grey70", "TRUE" = "#0072B2"),
                     guide = "none") +
  labs(title = "Q3 (detail): L. crispatus vs IL-8",
       x = "L. crispatus DNA", y = "IL-8 concentration")
```

**Interpretation:** There are several significant correlations, but they form
two interesting patterns:

1. **Negative correlations** (L. crispatus and L. iners with IL-1a, IL-1b,
   IL-6, IL-8) — more bacteria = less inflammation. This makes biological sense:
   a healthier microbiome may suppress inflammatory cytokines.
2. **Positive correlations** (both bacteria with IP-10, MIG, MIP-3a) — more
   bacteria = more of these chemokines. This might reflect immune surveillance
   (the immune system "noticing" bacteria).

Importantly, L. crispatus and L. iners show the **same direction** for every
cytokine (both negative or both positive) — suggesting the correlation is driven
by total bacterial load rather than a specific species effect.

---

# Q4: Participant Characteristics — Do age / birth control affect outcomes?

```{r}
#| fig.width: 8
#| fig.height: 5

qpcr |>
  left_join(select(sample_ids, sample_id, pid), by = "sample_id") |>
  left_join(select(metadata, pid, birth_control), by = "pid") |>
  pivot_longer(c(qpcr_bacteria, qpcr_crispatus, qpcr_iners),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome_label = case_when(
    outcome == "qpcr_bacteria"  ~ "Total bacteria",
    outcome == "qpcr_crispatus" ~ "L. crispatus",
    outcome == "qpcr_iners"     ~ "L. iners"
  )) |>
  ggboxplot(x = "birth_control", y = "value", color = "birth_control",
            palette = c("#D55E00", "#56B4E9"), add = "jitter") +
  facet_wrap(~ outcome_label, scales = "free_y") +
  stat_compare_means(method = "wilcox.test", label = "p.format") +
  labs(title = "Q4: Bacterial outcomes by birth control type",
       x = "Birth Control", y = "DNA copy count") +
  theme(legend.position = "none")
```

```{r}
#| echo: false
qpcr_bc <- qpcr |>
  left_join(select(sample_ids, sample_id, pid), by = "sample_id") |>
  left_join(select(metadata, pid, birth_control), by = "pid")

bc_pvals <- map_dbl(c("qpcr_bacteria", "qpcr_crispatus", "qpcr_iners"), ~
  wilcox.test(qpcr_bc[[.x]] ~ qpcr_bc$birth_control)$p.value) |> round(4)
```

**Birth control results:**

| Outcome | p-value | Significant? |
|---|---|---|
| Total bacteria | `r bc_pvals[1]` | `r ifelse(bc_pvals[1] < 0.05, "✅", "❌")` |
| *L. crispatus* | `r bc_pvals[2]` | `r ifelse(bc_pvals[2] < 0.05, "✅", "❌")` |
| *L. iners* | `r bc_pvals[3]` | `r ifelse(bc_pvals[3] < 0.05, "✅", "❌")` |

### Age: does age correlate with bacterial levels?

```{r}
#| fig.width: 8
#| fig.height: 4

qpcr |>
  left_join(select(sample_ids, sample_id, pid), by = "sample_id") |>
  left_join(select(metadata, pid, age), by = "pid") |>
  pivot_longer(c(qpcr_bacteria, qpcr_crispatus, qpcr_iners),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome_label = case_when(
    outcome == "qpcr_bacteria"  ~ "Total bacteria",
    outcome == "qpcr_crispatus" ~ "L. crispatus",
    outcome == "qpcr_iners"     ~ "L. iners"
  )) |>
  ggplot(aes(x = age, y = value)) +
  geom_point(alpha = 0.5, size = 2.5) +
  geom_smooth(method = "lm", color = "#0072B2", se = FALSE, linetype = "dashed") +
  stat_cor(method = "spearman", cor.coef.name = "rho") +
  facet_wrap(~ outcome_label, scales = "free_y") +
  labs(title = "Q4: Bacterial levels by age",
       x = "Age (years)", y = "DNA copy count")
```

**Interpretation:** Birth control type does not significantly predict any
bacterial outcome (all p > 0.05). Age also shows no strong correlations. This is
good news — it means the birth control imbalance between arms (which we saw
in Table 1, p = 0.017) is unlikely to be a major confounder for the
yogurt analyses.

---

# Q5: Combined — Does the yogurt effect depend on time point? (Interaction)

This is the most interesting question. Maybe yogurt doesn't matter at baseline
(when everyone's microbiome is intact), but it affects how the microbiome
**recovers** after antibiotics. That's a **time × arm interaction**.

```{r}
#| fig.width: 10
#| fig.height: 5

qpcr_long |>
  ggboxplot(x = "time_point", y = "value", color = "arm",
            palette = c("#E69F00", "#0072B2"), add = "jitter") +
  facet_wrap(~ outcome_label, scales = "free_y") +
  stat_compare_means(aes(group = arm), method = "wilcox.test",
                     label = "p.format", label.y.npc = "bottom") +
  labs(x = "Time Point", y = "DNA copy count",
       title = "Q5: Yogurt vs diet — split by time point",
       subtitle = "P-values compare yogurt vs diet WITHIN each time point",
       color = "Study Arm")
```

```{r}
#| echo: false
interaction_pvals <- qpcr_long |>
  group_by(outcome_label, time_point) |>
  summarise(p = wilcox.test(value ~ arm)$p.value, .groups = "drop") |>
  mutate(p = round(p, 4))
```

**P-values: yogurt vs diet at each time point:**

| Outcome | At baseline | After antibiotics |
|---|---|---|
`r paste0("| ", unique(interaction_pvals$outcome_label), " | ", round(interaction_pvals$p[interaction_pvals$time_point == "baseline"], 4), " | ", round(interaction_pvals$p[interaction_pvals$time_point == "after_antibiotic"], 4), " |", collapse = "\n")`

### Change scores: visualizing the trajectory

A cleaner way: compute `after_antibiotic - baseline` for each participant,
then compare the **amount of change** between yogurt and diet groups.

```{r}
#| fig.width: 8
#| fig.height: 5

qpcr_change <- qpcr_long |>
  select(pid, arm, outcome_label, time_point, value) |>
  distinct() |>
  pivot_wider(names_from = time_point, values_from = value) |>
  mutate(change = after_antibiotic - baseline) |>
  filter(!is.na(change))

ggboxplot(qpcr_change, x = "arm", y = "change", color = "arm",
          palette = c("#E69F00", "#0072B2"), add = "jitter",
          add.params = list(size = 2, alpha = 0.5)) +
  facet_wrap(~ outcome_label, scales = "free_y") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  stat_compare_means(comparisons = list(c("unchanged_diet", "yogurt")),
                     method = "wilcox.test", label = "p.format") +
  labs(x = "Study Arm", y = "Change (after – baseline)",
       title = "Q5: Change after antibiotics — does yogurt alter the trajectory?",
       subtitle = "Positive = increase. 0 = no change. Negative = decrease after antibiotics.") +
  theme(legend.position = "none")
```

```{r}
#| echo: false
delta_pvals <- qpcr_change |>
  group_by(outcome_label) |>
  summarise(p = wilcox.test(change ~ arm)$p.value, .groups = "drop") |>
  mutate(p = round(p, 4))
```

**P-values for change scores (yogurt vs diet):**

| Outcome | p-value | Significant? |
|---|---|---|
`r paste0("| ", delta_pvals$outcome_label, " | ", delta_pvals$p, " | ", ifelse(delta_pvals$p < 0.05, "✅ Yes", "❌ No (but trending)"), " |", collapse = "\n")`

**Interpretation: The interaction is not statistically significant at α =
0.05 for any outcome, but there are interesting trends.** In particular,
*L. crispatus* change scores have p = `r delta_pvals$p[2]` — a borderline
signal worth discussing. Looking at the split-by-time figure, at baseline the
two arms look nearly identical (p ≈ 0.98), but after antibiotics there's a
visible (though non-significant) separation (p ≈ 0.076). The trend suggests
yogurt eaters may retain more *L. crispatus* after antibiotics, but the study
is **underpowered** to detect this effect at the standard significance level.

---

## Summary: all findings in one table

| Question | Key finding | p-value(s) | Bottom line |
|---|---|---|---|
| Q1: Treatment (yogurt vs diet) | No difference in bacteria or cytokines | 0.11–0.61 | Yogurt alone doesn't shift microbiome |
| Q2: Time (baseline vs after abx) | **All bacteria change** dramatically | **< 0.001** | Antibiotics clearly disrupt microbiome |
| Q3: Bacteria × cytokine correlation | Several significant, two directions | 0.000–0.01 | Bacteria linked to immune markers |
| Q4: Participant traits (BC / age) | No strong effects | 0.09–0.81 | Confounders unlikely |
| Q5: Interaction (time × arm) | **Trend but not significant** | **0.07** (crispatus) | May be underpowered |

**Bottom line:** The strongest finding is that antibiotics profoundly change
the vaginal microbiome (Q2). There's a **suggestive trend** that yogurt may
help *L. crispatus* recover after antibiotics (Q5, p ≈ 0.07), but the study
may need more participants to confirm this. The correlations between bacteria
and cytokines (Q3) hint at a biological link between the microbiome and
immune activity.

```{r}
#| echo: false
cat("\n**Recommendation:** If this were a real study, the next step would be a\n",
    "larger trial focused specifically on the post-antibiotic recovery window,\n",
    "with L. crispatus as the primary outcome. The current data is consistent\n",
    "with a yogurt benefit but can't prove it.\n")
```