---
title: "Comparison Figures — Yogurt Study"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false
library(tidyverse)
library(ggpubr)
library(here)

# Read and join arm + time_point info
sample_ids <- read_csv(here("data", "group_a_yogurt", "00_sample_ids_yogurt.csv"),
                       show_col_types = FALSE)

qpcr <- read_csv(here("data", "group_a_yogurt", "02_qpcr_results_yogurt.csv"),
                 show_col_types = FALSE) |>
  left_join(sample_ids |> select(sample_id, arm), by = "sample_id")

luminex <- read_csv(here("data", "group_a_yogurt", "03_luminex_results_yogurt.csv"),
                    show_col_types = FALSE) |>
  left_join(sample_ids |> select(sample_id, arm), by = "sample_id")
```

## qPCR outcomes — faceted by outcome

Three bacterial DNA measures, split by study arm. Each panel shows one outcome,
and `stat_compare_means()` draws the Wilcoxon p-value on the plot.

```{r}
# Pivot longer so we can facet by outcome name
qpcr_long <- qpcr |>
  pivot_longer(
    cols = c(qpcr_bacteria, qpcr_crispatus, qpcr_iners),
    names_to = "outcome",
    values_to = "value"
  ) |>
  # Clean up labels for the facet strips
  mutate(outcome = case_when(
    outcome == "qpcr_bacteria"  ~ "Total bacteria",
    outcome == "qpcr_crispatus" ~ "L. crispatus",
    outcome == "qpcr_iners"     ~ "L. iners"
  ))
```

```{r}
ggpubr::ggboxplot(
  qpcr_long,
  x = "arm",
  y = "value",
  color = "arm",
  palette = c("#E69F00", "#0072B2"),
  add = "jitter",
  add.params = list(size = 1.5, alpha = 0.4)
) +
  facet_wrap(~ outcome, scales = "free_y") +
  stat_compare_means(comparisons = list(c("unchanged_diet", "yogurt")),
                     method = "wilcox.test",
                     label = "p.format") +
  labs(x = "Study Arm", y = "DNA copy count",
       title = "qPCR outcomes by study arm") +
  theme(legend.position = "none")
```

Looking at *L. crispatus* — notice how a few high-outlier samples stretch the
y-axis. On the log scale, the pattern is clearer:

```{r}
ggpubr::ggboxplot(
  qpcr_long |> mutate(value = value + 1),  # +1 so log(0) = log(1) = 0
  x = "arm",
  y = "value",
  color = "arm",
  palette = c("#E69F00", "#0072B2"),
  add = "jitter",
  add.params = list(size = 1.5, alpha = 0.4)
) +
  facet_wrap(~ outcome, scales = "free_y") +
  scale_y_log10(labels = scales::label_number()) +
  stat_compare_means(comparisons = list(c("unchanged_diet", "yogurt")),
                     method = "wilcox.test",
                     label = "p.format") +
  labs(x = "Study Arm", y = "DNA copy count (log scale)",
       title = "qPCR outcomes by study arm (log scale)") +
  theme(legend.position = "none")
```

---

## Cytokine immune markers — faceted by cytokine

All 10 Luminex cytokines. **Caution:** several cytokines have most values flagged
as "out of range" in the lab assay — those measurements may not be reliable.

```{r}
#| warning: true
# Data note: IL-10 is constant (0.767, all "out of range") — won't plot.

cytokines_valid <- c("IFN-Y", "TNFa", "IL-1a", "IL-1b", "IL-6", "IL-8", "IP-10", "MIG", "MIP-3a")

luminex |>
  filter(cytokine %in% cytokines_valid) |>
  ggboxplot(
    x = "arm",
    y = "conc",
    color = "arm",
    palette = c("#E69F00", "#0072B2"),
    add = "jitter",
    add.params = list(size = 1, alpha = 0.3)
  ) +
  facet_wrap(~ cytokine, scales = "free_y", nrow = 3) +
  stat_compare_means(comparisons = list(c("unchanged_diet", "yogurt")),
                     method = "wilcox.test",
                     label = "p.format",
                     tip.length = 0.01) +
  labs(x = "Study Arm", y = "Concentration",
       title = "Cytokine concentrations by study arm",
       caption = "Wilcoxon rank-sum p-values shown on each panel.") +
  theme(legend.position = "none")
```

Since cytokine data is often right-skewed, the log scale is more informative:

```{r}
luminex |>
  filter(cytokine %in% cytokines_valid, conc > 0) |>
  ggboxplot(
    x = "arm",
    y = "conc",
    color = "arm",
    palette = c("#E69F00", "#0072B2"),
    add = "jitter",
    add.params = list(size = 1, alpha = 0.3)
  ) +
  facet_wrap(~ cytokine, scales = "free_y", nrow = 3) +
  scale_y_log10(labels = scales::label_number()) +
  stat_compare_means(comparisons = list(c("unchanged_diet", "yogurt")),
                     method = "wilcox.test",
                     label = "p.format",
                     tip.length = 0.01) +
  labs(x = "Study Arm", y = "Concentration (log scale)",
       title = "Cytokine concentrations by study arm (log scale)",
       caption = "Wilcoxon rank-sum p-values. Zero concentrations excluded for log transform.") +
  theme(legend.position = "none")
```

---

## Data quality note

```{r}
#| echo: false
luminex |>
  group_by(cytokine, limits) |>
  summarise(n = n(), .groups = "drop") |>
  pivot_wider(names_from = limits, values_from = n, values_fill = 0) |>
  mutate(prop_unreliable = round(`out of range` / (`out of range` + `within limits`) * 100, 0)) |>
  knitr::kable(caption = "Cytokine data quality: how many measurements are 'out of range'?")
```

The dotted line on the log-scale plots still includes these "out of range" values;
be cautious with cytokines where >50% of data is flagged unreliable (IFN-Y, IL-10,
TNFa).

---

## Code summary

The pattern is always the same:

```r
library(ggpubr)

ggboxplot(data, x = "arm", y = "outcome",
          color = "arm", add = "jitter") +
  facet_wrap(~ group_var, scales = "free_y") +
  stat_compare_means(comparisons = list(c("group1", "group2")),
                     method = "wilcox.test",
                     label = "p.format")
```

- `ggboxplot()` — draws the boxplots with optional jittered points
- `facet_wrap()` — splits into multiple panels, one per outcome
- `stat_compare_means()` — adds the p-value; auto-chooses Wilcoxon for 2 groups
  or Kruskal-Wallis for 3+, or you can set `method = "t.test"`