---
title: "Table 1 — Yogurt Study Participant Characteristics"
format: html
---

```{r setup, include=FALSE}
library(tidyverse)
library(gtsummary)
```

```{r}
# Load participant metadata
metadata <- read_csv("../data/group_a_yogurt/01_participant_metadata_yogurt.csv")
```

## Variable-by-variable decisions

### `age`
- **Default summary:** median (IQR) — keeps the default. Age ranges from 24–37
  and is roughly symmetric, so median (IQR) is perfectly reasonable.
- **Default test:** Wilcoxon rank-sum test — **override to `t.test`**. Age is roughly
  normally distributed (bell-shaped), so a t-test is slightly more powerful here.
  Both tests will agree, but t-test matches the mean if we wanted that.

### `days_since_last_sex`
- **Default summary:** median (IQR) — **keep default**. This variable is heavily
  right-skewed (most values 1–14, outliers up to 48). Median (IQR) is the right
  choice for skewed data.
- **Default test:** Wilcoxon rank-sum test — **keep default**. Wilcoxon is
  non‑parametric and doesn't assume normality, which is ideal for skewed data.

### `birth_control`
- **Default summary:** n (%) — **keep default**. Two categories (Depoprovera vs
  no hormonal birth control), nicely balanced (25 vs 26).
- **Default test:** Chi-squared test — **keep default**. All expected counts are
  well above 5, so chi-squared is fine.

### `education`
- **Default summary:** n (%) — **keep default**. Four ordered categories, but in a
  standard Table 1 they're usually presented as counts with percentages.
- **Default test:** Chi-squared test — **override to `fisher.test`**. One of the
  categories (e.g., `less than grade 9`) has only 9 people total, so some cells
  may have small expected counts. Fisher's exact test is more reliable in that
  situation.

### `pid` (participant ID)
- **Exclude** from the table — it's an identifier, not a characteristic.

### `sex`
- **Exclude** — all 51 participants are female. A constant column adds no
  information and can't be tested.

---

## Build the Table 1

```{r}
table_one <- metadata |>
  select(-pid, -sex) |>                  # drop identifier and constant column
  tbl_summary(
    by = arm,                             # compare yogurt vs unchanged_diet
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 1,
      all_categorical() ~ c(0, 1)
    ),
    label = list(
      days_since_last_sex ~ "Days since last sex",
      birth_control ~ "Birth control method"
    ),
    missing_text = "Missing"
  ) |>
  add_p(
    test = list(
      all_continuous() ~ "wilcox.test",
      age ~ "t.test",                      # override: use t-test for age
      education ~ "fisher.test"            # override: use Fisher's for education
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 2)
  ) |>
  add_overall() |>                         # add an "Overall" column
  bold_labels() |>
  modify_header(label = "**Characteristic**")

table_one
```

> **Note on the test overrides:** The `all_continuous() ~ "wilcox.test"` sets
> the baseline for all continuous variables, and then the specific overrides
> (`age ~ "t.test"`) take precedence for that variable.

---

## What the table tells us

```{r}
# Optional: print with p-values highlighted
table_one |>
  modify_footnote(
    all_stat_cols() ~ "Median (IQR); n (%)"
  )
```
