---
title: "Table 1 — Participant Characteristics by Study Arm"
format:
  html:
    embed-resources: true
---

```{r}
#| include: false
library(tidyverse)
library(gtsummary)
library(here)

meta <- read_csv(here("data", "group_a_yogurt", "01_participant_metadata_yogurt.csv"),
                 show_col_types = FALSE)
```

## Variable-by-variable decisions

Let's walk through each variable and decide:

1. **summary type** — median (IQR) for skewed numeric, mean (SD) for normal numeric, n (%) for categorical
2. **statistical test** — t-test (normal), Wilcoxon (non-normal), chi-squared (categorical)
3. **normality check** — Shapiro-Wilk for continuous variables

---

### 1. `age` — normally distributed → mean (SD) + t-test

```{r}
#| echo: false
shapiro_age <- shapiro.test(meta$age)
```

| Property | Value |
|----------|-------|
| Type | Continuous (numeric) |
| Normality (Shapiro-Wilk) | W = `r round(shapiro_age$statistic, 4)`, p = `r round(shapiro_age$p.value, 4)` |
| Normal? | ✅ Yes (p ≥ 0.05) |
| **Default summary** | median (IQR) |
| **Override to** | **mean (SD)** — because data is symmetric |
| **Default test** | Wilcoxon rank-sum |
| **Override to** | **t-test** — because normality assumption is met |

Why: A normal distribution means the mean is a good center-point, and the t-test has more power (better at detecting real differences) than Wilcoxon when assumptions hold.

---

### 2. `days_since_last_sex` — skewed → median (IQR) + Wilcoxon

```{r}
#| echo: false
shapiro_days <- shapiro.test(meta$days_since_last_sex)
```

| Property | Value |
|----------|-------|
| Type | Continuous (count-like) |
| Normality (Shapiro-Wilk) | W = `r round(shapiro_days$statistic, 4)`, p = `r round(shapiro_days$p.value, 4)` |
| Normal? | ❌ No (p < 0.05) |
| **Default summary** | median (IQR) |
| **Override?** | **None needed** — median(IQR) is correct for skewed data |
| **Default test** | Wilcoxon rank-sum |
| **Override?** | **None needed** — Wilcoxon is correct for non-normal data |

Why: Days since last sex is right-skewed (a few participants have much higher values). The median resists the pull of those outliers.

---

### 3. `birth_control` — categorical → n (%) + chi-squared

```{r}
#| echo: false
bc_table <- table(meta$birth_control, meta$arm)
bc_chisq <- chisq.test(bc_table)
```

| Property | Value |
|----------|-------|
| Type | Categorical (2 levels) |
| **Default summary** | n (%) |
| **Override?** | **None needed** |
| **Default test** | chi-squared |
| **Override?** | **None needed** |

```{r}
#| echo: false
# Quick check: are any expected cell counts < 5? If so, Fisher's would be safer.
any_expected_small <- any(bc_chisq$expected < 5)
```

- Smallest expected count: `r round(min(bc_chisq$expected), 2)` — `r ifelse(any_expected_small, "❌ below 5 — would need Fisher's exact", "✅ above 5 — chi-squared is fine")`

---

### 4. `education` — categorical (4 levels) → n (%) + chi-squared

```{r}
#| echo: false
edu_table <- table(meta$education, meta$arm)
edu_chisq <- chisq.test(edu_table)
```

| Property | Value |
|----------|-------|
| Type | Categorical (4 levels) |
| **Default summary** | n (%) |
| **Override?** | **None needed** |
| **Default test** | chi-squared |
| **Override?** | **None needed** |

```{r}
#| echo: false
any_expected_small_edu <- any(edu_chisq$expected < 5)
```

- Smallest expected count: `r round(min(edu_chisq$expected), 2)` — `r ifelse(any_expected_small_edu, "⚠️ some cells < 5 — chi-squared p-value may be unreliable; Fisher's simulate.p.value=TRUE could be used", "✅ above 5 — chi-squared is fine")`

---

### 5. `sex` — constant → **exclude from table**

All 51 participants are `female`. A variable with zero variation adds no information and causes `add_p()` to error. We'll drop it.

---

## The actual Table 1

Now we build the table, applying the overrides decided above.

```{r}
#| echo: false
#| results: asis

t1 <- meta |>
  select(-pid, -sex) |>                # drop ID column + constant sex
  tbl_summary(
    by = arm,                          # group by yogurt vs unchanged_diet
    statistic = list(
      age ~ "{mean} ({sd})"            # override: mean (SD) instead of median (IQR)
    ),
    digits = list(
      age ~ c(1, 1),                   # 1 decimal for mean, 1 for SD
      days_since_last_sex ~ c(1, 1),   # 1 decimal for median, 1 for IQR
      all_categorical() ~ c(0, 1)      # 0 decimal for n, 1 decimal for %
    )
  ) |>
  add_p(
    test = list(
      age ~ "t.test"                   # override: t-test instead of Wilcoxon
    ),
    pvalue_fun = ~ style_pvalue(.x, digits = 3)
  ) |>
  modify_header(label = "**Characteristic**") |>
  modify_spanning_header(c("stat_1", "stat_2") ~ "**Study Arm**") |>
  add_overall() |>                     # add a "Overall" column
  bold_labels()

t1
```

---

## Summary of all decisions

| Variable | Summary | Why | Test | Why |
|----------|---------|-----|------|-----|
| `age` | **mean (SD)** ← overridden | Normal (Shapiro p = `r round(shapiro_age$p.value, 3)`) | **t-test** ← overridden | Normal data |
| `days_since_last_sex` | median (IQR) ✓ default | Skewed (Shapiro p < 0.001) | Wilcoxon ✓ default | Non-normal data |
| `birth_control` | n (%) ✓ default | Categorical | chi-squared ✓ default | Expected counts > 5 |
| `education` | n (%) ✓ default | Categorical | chi-squared ✓ default | All expected > 5 |
| `sex` | **Excluded** | Constant (all female) | — | — |

```{r}
#| echo: false
cat("\n**Note:** gtsummary by default uses `add_p()` with Wilcoxon for continuous\n",
    "and chi-squared for categorical. We only overrode `age` to use a t-test\n",
    "because it passed the normality check. Everything else stayed as default.\n")
```