---
title: "Hello, R!"
format: html
---

Welcome! This is your very first notebook. If you can see a chart appear when
you run the code below, your environment is working and you are ready to go.
You do not need to understand every line yet — that is what the workshop is
for. For now, just press play and watch what happens.

```{r}
library(tidyverse)
library(palmerpenguins)

ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  labs(
    title = "Penguin flipper length vs. body mass",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species"
  )
```

```{r}
library(tidyverse)
sample_ids <- read_csv("data/group_a_yogurt/00_sample_ids_yogurt.csv")
metadata <- read_csv("data/group_a_yogurt/01_participant_metadata_yogurt.csv")
qpcr <- read_csv("data/group_a_yogurt/02_qpcr_results_yogurt.csv")
luminex <- read_csv("data/group_a_yogurt/03_luminex_results_yogurt.csv")
```
```{r}
glimpse(sample_ids)
glimpse(metadata)
glimpse(qpcr)
glimpse(luminex)
count(sample_ids, arm)
count(sample_ids, time_point)
count(luminex, cytokine)
summary(metadata$age)
qpcr_joined <- qpcr |>
  left_join(sample_ids, by = "sample_id")

glimpse(qpcr_joined)
luminex_joined <- luminex |>
  left_join(sample_ids, by = "sample_id")

glimpse(luminex_joined)
ggplot(qpcr_joined, aes(x = arm, y = qpcr_bacteria, fill = arm )) +
  geom_boxplot() +
  labs(title = "Total bacteria by study arm")
ggplot(qpcr_joined, aes(x = arm, y = qpcr_crispatus, fill = arm )) +
  geom_boxplot() +
  labs(title = "L. crispatus DNA by study arm")
luminex_joined |>
  filter(cytokine == "IL-1a") |>
  ggplot(aes(x = arm, y = conc, fill = arm )) +
  geom_boxplot() +
  labs(title = "IL-1a concentration by study arm")

```
is your own? Open the terminal and ask Pi to help —
for example, "Can you change the plot to show bill length instead of flipper
length?" Experimenting is the best way to learn.
