---
title: "Module 2 — Intro to R and Quarto"
format: html
editor: source
execute:
  warning: false
  message: false
---

This is your workspace for the Module 2 worksheet. The full instructions live
on the lesson page — keep it open alongside this notebook. You will build up
this notebook one part at a time: fill in the empty chunk under each heading,
press **Run Cell** to see what happens, and write a sentence of your own
between the chunks. Render with **Preview** whenever the worksheet tells you to.

Always load the packages you need in a setup chunk at the top. Run this first,
every time you open the notebook:

```{r setup}
library(tidyverse)
library(here)
```

## Part 1 — Run a chunk, write a sentence, render

```{r}
# 3+5+10
10 * (5+1)
2 ** 10
```

## Part 2 — Read the data and look at it

```{r}
# 64 / 2 + 2
64 / (2 + 2)
```

## Part 3 — Pull out columns and summarise them

```{r}
table_02 <- read_csv(here("data/instructional_dataset/02_visit_clinical_measurements_UKZN_workshop_2023.csv"))

table_02
glimpse(table_02)
table_02$nugent_score
mean(table_02$crp_blood, na.rm = TRUE)
crp_mean <- mean(table_02$crp_blood, na.rm = TRUE)
crp_mean
```

## Part 4 — Build something with Pi

```{r}
table_02 %>%
group_by(arm) %>%
mean(table_02$nugent_score, na.rm = TRUE)
table_02 %>%                                                                                          
     group_by(arm) %>%                                                                                   
     summarise(                                                                                          
       mean_nugent = mean(nugent_score, na.rm = TRUE),                                                   
       mean_crp   = mean(crp_blood, na.rm = TRUE),                                                       
       mean_ph    = mean(ph, na.rm = TRUE)                                                               
     )              
select(-pid, -time_point) %>% tbl_summary(by = arm)                                                                       
                                   
```

## Capstone — make it your own

```{r}
# your code here
```

## Before you go

Pick one step today that confused you. Ask Pi to explain it, then push back:
*"why did you use X instead of Y?"* Notice when Pi sounds confident versus when
it is guessing.
