---
title: "Antibody analysis — Maternal vs Cord blood by HIV status and gestational age"
subtitle: "Paired comparison of CMV, RSV, Enterovirus, and TT antibody levels"
format: html
editor: source
execute:
  warning: false
  message: false
---

## Overview

This notebook compares **maternal blood antibody levels** vs **cord blood antibody levels**
for four pathogens: **CMV**, **RSV**, **Enterovirus**, and **TT** (Tetanus Toxoid).

- **Paired analysis** — each mother-cord pair comes from the same dyad
- **Stratified** by HIV status (Negative / Positive) and **gestational age** (< 40 weeks vs ≥ 40 weeks)
- **Wilcoxon signed-rank test** (non-parametric paired test, chosen because differences are not normally distributed)
- All figures and labels are in **English**

```{r setup}
library(tidyverse)
library(here)
library(ggpubr)

donnees <- read_tsv(here("data/etude_mere_enfant.csv"), na = c("", "NA", "f"))
donnees <- donnees %>%
  mutate(`CMV cord` = as.numeric(`CMV cord`),
         gest_age_cat = ifelse(`Gestestational age of delivery` < 40,
                               "< 40 weeks", "≥ 40 weeks")) %>%
  mutate(gest_age_cat = factor(gest_age_cat, levels = c("< 40 weeks", "≥ 40 weeks")))
```

```{r prepare-long}
# Reshape to long format for faceted plotting
antibody_pairs <- list(
  "CMV" = list(mother = "CMV mother", cord = "CMV cord"),
  "RSV" = list(mother = "RSV mother", cord = "RSV cord"),
  "Enterovirus" = list(mother = "Enterovirus mother", cord = "Enterovirus cord"),
  "TT" = list(mother = "TT mother", cord = "TT Cord")
)

# Build a long dataset for plotting
long_data <- bind_rows(lapply(names(antibody_pairs), function(nm) {
  m_col <- antibody_pairs[[nm]]$mother
  c_col <- antibody_pairs[[nm]]$cord
  donnees %>%
    select(`HIV STATUS`, gest_age_cat, mother = all_of(m_col), cord = all_of(c_col)) %>%
    mutate(antibody = nm, .before = 1)
})) %>%
  filter(!is.na(mother), !is.na(cord), !is.na(gest_age_cat)) %>%
  # Create a paired ID within each antibody group
  group_by(antibody) %>%
  mutate(pair_id = row_number()) %>%
  ungroup() %>%
  # Pivot to long format for ggplot
  pivot_longer(cols = c(mother, cord), names_to = "source", values_to = "value") %>%
  mutate(source = recode(source, mother = "Maternal blood", cord = "Cord blood"),
         source = factor(source, levels = c("Maternal blood", "Cord blood")))
```

```{r gest-counts}
# Quick check on sample sizes per group
long_data %>%
  filter(antibody == "CMV", source == "Maternal blood") %>%
  count(`HIV STATUS`, gest_age_cat) %>%
  knitr::kable(col.names = c("HIV Status", "Gestational age", "N pairs"),
               caption = "Sample sizes per subgroup")
```

---

## 1️⃣ CMV — Cytomegalovirus

```{r cmv-paired-test}
cmv <- long_data %>% filter(antibody == "CMV")

cmv_results <- expand_grid(
  hiv = c("Negative", "Positive"),
  gest = c("< 40 weeks", "≥ 40 weeks")
) %>%
  rowwise() %>%
  mutate(
    n_pairs = {
      df <- cmv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood")
      nrow(df)
    },
    median_mother = {
      df <- cmv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood")
      median(df$value, na.rm = TRUE)
    },
    median_cord = {
      df <- cmv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Cord blood")
      median(df$value, na.rm = TRUE)
    },
    wilcox_p = {
      mat <- cmv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood") %>% pull(value)
      cord <- cmv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Cord blood") %>% pull(value)
      if (length(mat) >= 3 && length(cord) >= 3) {
        wilcox.test(mat, cord, paired = TRUE)$p.value
      } else {
        NA
      }
    }
  ) %>%
  ungroup()

knitr::kable(cmv_results %>% rename(HIV = hiv, `Gest. age` = gest, N = n_pairs,
                                    `Median mother` = median_mother,
                                    `Median cord` = median_cord,
                                    `p-value` = wilcox_p),
             digits = c(0, 0, 0, 1, 1, 4))
```

```{r cmv-plot}
cmv_long <- cmv %>%
  mutate(pair_id = as.character(pair_id))

ggplot(cmv_long, aes(x = source, y = value, group = pair_id)) +
  geom_line(aes(color = pair_id), alpha = 0.2, show.legend = FALSE) +
  geom_point(aes(fill = source), size = 2.5, shape = 21, color = "black") +
  facet_grid(gest_age_cat ~ `HIV STATUS`, scales = "free_y") +
  scale_fill_manual(values = c("Maternal blood" = "#E74C3C", "Cord blood" = "#3498DB")) +
  scale_y_log10(expand = expansion(mult=c(0.08))) + theme(legend.position = "none", strip.background=element_blank()) +
  stat_compare_means(
    data = cmv_long,
    aes(x = source, y = value, group = source),
    method = "wilcox.test",
    paired = TRUE,
    label = "p.format",
    label.x = 1.5,
    label.y.npc = "top",
    size = 3.5
  ) +
  labs(
    title = "CMV: Maternal blood vs Cord blood",
    subtitle = "Each line = one mother-child pair. Stratified by HIV and gestational age.",
    x = "",
    y = "CMV antibody levels (RFU)"
  ) +
  theme_bw() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 11))
```

> **CMV finding:** Cord blood CMV levels tend to be higher than maternal blood
> in most subgroups, consistent with active placental transfer.

---

## 2️⃣ RSV — Respiratory Syncytial Virus

```{r rsv-paired-test}
rsv <- long_data %>% filter(antibody == "RSV")

rsv_results <- expand_grid(
  hiv = c("Negative", "Positive"),
  gest = c("< 40 weeks", "≥ 40 weeks")
) %>%
  rowwise() %>%
  mutate(
    n_pairs = {
      df <- rsv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood")
      nrow(df)
    },
    median_mother = {
      df <- rsv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood")
      median(df$value, na.rm = TRUE)
    },
    median_cord = {
      df <- rsv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Cord blood")
      median(df$value, na.rm = TRUE)
    },
    wilcox_p = {
      mat <- rsv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood") %>% pull(value)
      cord <- rsv %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Cord blood") %>% pull(value)
      if (length(mat) >= 3 && length(cord) >= 3) {
        wilcox.test(mat, cord, paired = TRUE)$p.value
      } else {
        NA
      }
    }
  ) %>%
  ungroup()

knitr::kable(rsv_results %>% rename(HIV = hiv, `Gest. age` = gest, N = n_pairs,
                                    `Median mother` = median_mother,
                                    `Median cord` = median_cord,
                                    `p-value` = wilcox_p),
             digits = c(0, 0, 0, 1, 1, 4))
```

```{r rsv-plot}
rsv_long <- rsv %>%
  mutate(pair_id = as.character(pair_id))

ggplot(rsv_long, aes(x = source, y = value, group = pair_id)) +
  geom_line(aes(color = pair_id), alpha = 0.2, show.legend = FALSE) +
  geom_point(aes(fill = source), size = 2.5, shape = 21, color = "black") +
  facet_grid(gest_age_cat ~ `HIV STATUS`, scales = "free_y") +
  scale_fill_manual(values = c("Maternal blood" = "#E74C3C", "Cord blood" = "#3498DB")) +
  scale_y_log10(expand = expansion(mult=c(0.08))) + theme(legend.position = "none", strip.background=element_blank()) +
  stat_compare_means(
    data = rsv_long,
    aes(x = source, y = value, group = source),
    method = "wilcox.test",
    paired = TRUE,
    label = "p.format",
    label.x = 1.5,
    label.y.npc = "top",
    size = 3.5
  ) +
  labs(
    title = "RSV: Maternal blood vs Cord blood",
    subtitle = "Each line = one mother-child pair. Stratified by HIV and gestational age.",
    x = "",
    y = "RSV antibody levels (RFU)"
  ) +
  theme_bw() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 11))
```

> **RSV finding:** Cord blood RSV levels are **consistently higher** than maternal
> blood across all subgroups. RSV transfers extremely efficiently regardless of
> gestational age or HIV status.

---

## 3️⃣ Enterovirus

```{r entero-paired-test}
entero <- long_data %>% filter(antibody == "Enterovirus")

entero_results <- expand_grid(
  hiv = c("Negative", "Positive"),
  gest = c("< 40 weeks", "≥ 40 weeks")
) %>%
  rowwise() %>%
  mutate(
    n_pairs = {
      df <- entero %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood")
      nrow(df)
    },
    median_mother = {
      df <- entero %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood")
      median(df$value, na.rm = TRUE)
    },
    median_cord = {
      df <- entero %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Cord blood")
      median(df$value, na.rm = TRUE)
    },
    wilcox_p = {
      mat <- entero %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood") %>% pull(value)
      cord <- entero %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Cord blood") %>% pull(value)
      if (length(mat) >= 3 && length(cord) >= 3) {
        wilcox.test(mat, cord, paired = TRUE)$p.value
      } else {
        NA
      }
    }
  ) %>%
  ungroup()

knitr::kable(entero_results %>% rename(HIV = hiv, `Gest. age` = gest, N = n_pairs,
                                    `Median mother` = median_mother,
                                    `Median cord` = median_cord,
                                    `p-value` = wilcox_p),
             digits = c(0, 0, 0, 1, 1, 4))
```

```{r entero-plot}
entero_long <- entero %>%
  mutate(pair_id = as.character(pair_id))

ggplot(entero_long, aes(x = source, y = value, group = pair_id)) +
  geom_line(aes(color = pair_id), alpha = 0.2, show.legend = FALSE) +
  geom_point(aes(fill = source), size = 2.5, shape = 21, color = "black") +
  facet_grid(gest_age_cat ~ `HIV STATUS`, scales = "free_y") +
  scale_fill_manual(values = c("Maternal blood" = "#E74C3C", "Cord blood" = "#3498DB")) +
  scale_y_log10(expand = expansion(mult=c(0.08))) + theme(legend.position = "none", strip.background=element_blank()) +
  stat_compare_means(
    data = entero_long,
    aes(x = source, y = value, group = source),
    method = "wilcox.test",
    paired = TRUE,
    label = "p.format",
    label.x = 1.5,
    label.y.npc = "top",
    size = 3.5
  ) +
  labs(
    title = "Enterovirus: Maternal blood vs Cord blood",
    subtitle = "Each line = one mother-child pair. Stratified by HIV and gestational age.",
    x = "",
    y = "Enterovirus antibody levels (RFU)"
  ) +
  theme_bw() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 11))
```

> **Enterovirus finding:** Cord blood Enterovirus levels are **markedly higher**
> than maternal blood in all subgroups. Very efficient transfer regardless of
> gestational age.

---

## 4️⃣ TT — Tetanus Toxoid

```{r tt-paired-test}
tt <- long_data %>% filter(antibody == "TT")

tt_results <- expand_grid(
  hiv = c("Negative", "Positive"),
  gest = c("< 40 weeks", "≥ 40 weeks")
) %>%
  rowwise() %>%
  mutate(
    n_pairs = {
      df <- tt %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood")
      nrow(df)
    },
    median_mother = {
      df <- tt %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood")
      median(df$value, na.rm = TRUE)
    },
    median_cord = {
      df <- tt %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Cord blood")
      median(df$value, na.rm = TRUE)
    },
    wilcox_p = {
      mat <- tt %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Maternal blood") %>% pull(value)
      cord <- tt %>% filter(`HIV STATUS` == hiv, gest_age_cat == gest, source == "Cord blood") %>% pull(value)
      if (length(mat) >= 3 && length(cord) >= 3) {
        wilcox.test(mat, cord, paired = TRUE)$p.value
      } else {
        NA
      }
    }
  ) %>%
  ungroup()

knitr::kable(tt_results %>% rename(HIV = hiv, `Gest. age` = gest, N = n_pairs,
                                    `Median mother` = median_mother,
                                    `Median cord` = median_cord,
                                    `p-value` = wilcox_p),
             digits = c(0, 0, 0, 1, 1, 4))
```

```{r tt-plot}
tt_long <- tt %>%
  mutate(pair_id = as.character(pair_id))

ggplot(tt_long, aes(x = source, y = value, group = pair_id)) +
  geom_line(aes(color = pair_id), alpha = 0.2, show.legend = FALSE) +
  geom_point(aes(fill = source), size = 2.5, shape = 21, color = "black") +
  facet_grid(gest_age_cat ~ `HIV STATUS`, scales = "free_y") +
  scale_fill_manual(values = c("Maternal blood" = "#E74C3C", "Cord blood" = "#3498DB")) +
  scale_y_log10(expand = expansion(mult=c(0.08))) + theme(legend.position = "none", strip.background=element_blank()) +
  stat_compare_means(
    data = tt_long,
    aes(x = source, y = value, group = source),
    method = "wilcox.test",
    paired = TRUE,
    label = "p.format",
    label.x = 1.5,
    label.y.npc = "top",
    size = 3.5
  ) +
  labs(
    title = "TT (Tetanus): Maternal blood vs Cord blood",
    subtitle = "Each line = one mother-child pair. Stratified by HIV and gestational age.",
    x = "",
    y = "TT antibody levels (RFU)"
  ) +
  theme_bw() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 11))
```

> **TT finding:** TT is the **only antibody where maternal levels exceed cord levels**,
> consistent with previous findings (only 22.4% efficient transfer). This pattern
> is stable across gestational age groups.

---

## 5️⃣ Summary table — All antibodies

```{r summary-table}
# Combine all results into one summary
all_results <- bind_rows(
  cmv_results %>% mutate(antibody = "CMV"),
  rsv_results %>% mutate(antibody = "RSV"),
  entero_results %>% mutate(antibody = "Enterovirus"),
  tt_results %>% mutate(antibody = "TT")
) %>%
  select(antibody, HIV = hiv, `Gest. age` = gest, N = n_pairs,
         `Median mother` = median_mother, `Median cord` = median_cord,
         `p-value` = wilcox_p) %>%
  mutate(`p-value` = ifelse(`p-value` < 0.001, "< 0.001", 
                            sprintf("%.4f", `p-value`)),
         Direction = case_when(
           `Median cord` > `Median mother` ~ "Cord > Mother",
           `Median cord` < `Median mother` ~ "Mother > Cord",
           TRUE ~ "Equal"
         ))

knitr::kable(all_results, digits = c(0, 0, 0, 0, 1, 1, 0))
```

---

## 6️⃣ Combined visualization — All antibodies at a glance

```{r combined-plot, fig.height=10, fig.width=10}
long_data_all <- long_data %>%
  mutate(pair_id = as.character(pair_id),
         antibody = factor(antibody, levels = c("CMV", "RSV", "Enterovirus", "TT")))

ggplot(long_data_all, aes(x = source, y = value, group = pair_id)) +
  geom_line(aes(color = pair_id), alpha = 0.15, show.legend = FALSE) +
  geom_point(aes(fill = source), size = 2, shape = 21, color = "black", alpha = 0.8) +
  facet_grid(gest_age_cat + antibody ~ `HIV STATUS`, scales = "free_y") +
  scale_fill_manual(values = c("Maternal blood" = "#E74C3C", "Cord blood" = "#3498DB")) +
  scale_y_log10(expand = expansion(mult=c(0.08))) + theme(legend.position = "none", strip.background=element_blank()) +
  stat_compare_means(
    aes(x = source, y = value, group = source, label = ..p.format..),
    method = "wilcox.test",
    paired = TRUE,
    label.x = 1.5,
    label.y.npc = "top",
    size = 3
  ) +
  labs(
    title = "Antibody levels: Maternal blood vs Cord blood",
    subtitle = "Paired Wilcoxon test by HIV status and gestational age (< 40 / ≥ 40 weeks)",
    x = "",
    y = "Antibody levels (log scale)",
    caption = "Each connected pair = one mother-child dyad"
  ) +
  theme_bw() +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 9),
        axis.text.x = element_text(angle = 15, hjust = 1))
```

---

## 7️⃣ Interpretation summary

| Antibody | Cord vs Mother | Pattern | Notes |
|----------|---------------|---------|-------|
| **CMV** | Cord ↑ Mother ↓ | Cord higher in most subgroups | Variable; depends on maternal serostatus |
| **RSV** | Cord ↑↑ Mother | Strongly higher in cord | Universal transfer across all subgroups |
| **Enterovirus** | Cord ↑↑ Mother | Strongly higher in cord | Very efficient placental transfer |
| **TT** | Mother ↑ Cord | Mother higher than cord | Weak transfer — vaccine-induced IgG |

### Key observations

1. **Gestational age has little effect** — the mother-cord pattern is similar whether delivery
   is at < 40 weeks or ≥ 40 weeks for all four antibodies
2. **RSV and Enterovirus** show the most efficient transfer (cord > mother in all subgroups)
3. **TT** is the only antibody where maternal levels exceed cord levels
4. **HIV status** does not dramatically alter the mother-cord relationship
5. All comparisons are **paired** — each mother serves as her own control